\bf Visual diagnostics for constrained optimisation with application to guided tours

### 2020-10-09 paper/paper.tex
include the plots for comparing diff seeds
remove codes, comments, list of plot elements that are unncessary, kol_1d_2var_beter_sim2 data

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

{[}Are we using projection pursuit/guided tour to better understand the convergence of optimization algorithms visually in combination with the algorithms discussed below? Or we are focusing on the optimisation problem only within the projection pursuit context? Some of the problems listed below are also applicable to optimization problem in general too. ppp{]}

@@

{\centering \includegraphics[width=0.6\linewidth,height=0.4\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/tour_path_keynote/tour_path_keynote.001}

@@

{[}outside the paper: I find the notation of current/target basis is confusing because the target basis in round \(j\) becomes the current basis in round \(j+1\). Also, when we start to have polish, the target basis may not be the current basis in the next round\ldots{} The place where current/target is most appropriate is probably when describing the interpolation where the first one is always the current basis and the last is always the target basis. I think it is better to leave this language in the code{]}

@@

Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that provides five points summary of the data, however, there are two pieces of information missing from the boxplot: 1) It does not report the number of points, and 2) the position of the last basis. This could be remedied by adding more layers using the concept of grammar of graphics. A label geometry is added at the bottom of the plot to show the number of points in each iteration and a line geometry links the last basis in each iteration. Further, an option to switch between displaying points and boxplot geometry is helpful since point geometry can be more intuitive for the iteration with few observations. This is achieved via a \texttt{cutoff} parameter. Below we define the searching point plot based on four different component layers

\begin{itemize}
\tightlist
\item
  Layer 1: boxplot geom

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the groups that have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value: \(Q_{I^{\prime}_t}(q)\) is mapped to the y-axis where \(Q_X(q)\), \(q = 0, 25, 50, 75, 100\) finds the qth-quantile of \(X\) and \(I^{\prime}_t\) denotes the index value of all the searching bases defined in Matrix \ref{eq:data-structure}.
  \end{itemize}
\item
  Layer 2: point geom

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that have count less than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\item
  Layer 3: line geom

  \begin{itemize}
  \tightlist
  \item
    data: filter the points with the highest index value in group \(j\)
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\item
  Layer 4: label geom

  \begin{itemize}
  \tightlist
  \item
    data: A count table of the number of observation in each iteration
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the y-axis for each label is \(0.99 * \text{MIN}\), where \(\text{MIN}\) is the smallest index value in the whole data object
  \item
    label: the number of observation is mapped to the label
  \end{itemize}
\end{itemize}

@@

Sometimes, we may be interested in exploring the points on the interpolation path since these are the points that will be played by the tour animation. The plot definition is as follows:

\begin{itemize}
\tightlist
\item
  Layer 1: point geom

  \begin{itemize}
  \tightlist
  \item
    data: filter the observations with \(S = \text{interpolation}\) and mutate \(t\) to be the row number of the subsetted tibble
  \item
    x: \(t\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\item
  Layer 2: line geom

  \begin{itemize}
  \tightlist
  \item
    using line geometry for the same data and aesthetics
  \end{itemize}
\end{itemize}

Figure \ref{toy-interp} presents the interpolation of three different tour paths. The upper plot shows a desirable interpolation in each iteration with the index value being progressively and monotonically increasing. While in the middle plot, the increases towards the target basis is not monotonical in the last two iterations and interpolated basis with higher index value can be found on the tour path. The lower plot is constructed using \texttt{search\_better\_random}, where a basis with smaller index value has a probabilistic chance of being accepted and we can observe a much more involved pattern. In iteration three, the probabilistic acceptance produces a monotonically decreasing interpolation whilst in iteration five, six and seven, with also a probabilistic acceptance, the interpolation is now non-monotonical. When the target basis has a higher index value, iteration eight reaches this basis by first decreasing the index value. While the interpolation plot shows different possibilities of the change in index value on the tour path, the diagnosis of the validity of each pattern and the related optimisation algorithms will be postponed to Section \ref{application}.

@@

~

@@

The visualisation can thus be defined as

\begin{itemize}
\tightlist
\item
  Layer 1: point geom

  \begin{itemize}
  \tightlist
  \item
    data: subset the basis of interest and arrange into a matrix format; perform PCA on the basis matrix and compute the projected basis on the first two principal components; bind the variables from the original global object and form a tibble
  \item
    x: the projected basis on the first principal component
  \item
    y: the projected basis on the second principal component
  \item
    colour: \(V\) is mapped to the colour aesthetic
  \end{itemize}
\end{itemize}

@@

Animating the plots introduced above is useful, especially in the case of PCA polot since it shows the bases found by the optimiser in its natural order.

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-1-1}

\caption{\label{toy-pca-aniamted}A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

Once the final basis has been found by an algorithm, one may want to push further to investigate whether there's an even better basis in the close neighbourhood. This motivates the polish search where the final basis is supplied to a new guided tour to search for any local breakthrough. The following code demonstrates the usage of \texttt{search\_polish}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tourr)}
\KeywordTok{library}\NormalTok{(ferrn)}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123456}\NormalTok{)}
\NormalTok{holes_2d_geo <-}\StringTok{ }
\StringTok{  }\KeywordTok{animate_xy}\NormalTok{(boa6,}
             \DataTypeTok{tour_path =} \KeywordTok{guided_tour}\NormalTok{(}\KeywordTok{holes}\NormalTok{(), }\DataTypeTok{d =} \DecValTok{2}\NormalTok{, }
                                     \DataTypeTok{search_f =}\NormalTok{ search_geodesic),}
             \DataTypeTok{rescale =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{last_basis <-}\StringTok{ }\KeywordTok{get_best}\NormalTok{(holes_2d_geo) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{pull}\NormalTok{(basis) }\OperatorTok{%>%}\StringTok{ }\NormalTok{.[[}\DecValTok{1}\NormalTok{]]}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123456}\NormalTok{)}
\NormalTok{holes_2d_geo_polish <-}\StringTok{ }
\StringTok{  }\KeywordTok{animate_xy}\NormalTok{(boa6, }
             \DataTypeTok{tour_path =} \KeywordTok{guided_tour}\NormalTok{(}\KeywordTok{holes}\NormalTok{(), }\DataTypeTok{d =} \DecValTok{2}\NormalTok{, }
                                     \DataTypeTok{search_f =}\NormalTok{ search_polish),}
             \DataTypeTok{rescale =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{start =}\NormalTok{ last_basis)}
\end{Highlighting}
\end{Shaded}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/kol-pca-1.pdf}
\caption{\label{fig:kol-pca}\label{kol-pca}Comparison between search\_better\_random and two different search\_better. In the trace plot, the two dashed horizontal lines are the index values of the two theoretical bases. serach\_better\_random and one of the serach\_better have found the global maximum but the other fails. The lower panel shows the PCA plot for each of the three algorithms. The two yellow dots, again, correspond to the theoretical bases with the one on the right being the global maximum. The paths of two search\_better are relatively short and simple while the search\_better\_random has a more sophisticated path. This is due to the fact that the interpolation at each iteration are longer, which shows its ability to find new bases at a distance.}
\end{figure}

@@

(further notice on the implementation in the tour package will be discussed in the tour developer meeting.)
The implementation of this projection has been divided into two pacakges, where the data collection happens when a tour is running and hence implemented in the existing \texttt{tourr} package. The diagnostics of the algorithm via plots and other auxiliary functions have been implemented in a new pacakge, \texttt{ferrn}.

@@

\begin{verbatim}
## # A tibble: 10 x 8
##       id basis             index_val info         tries  loop method       alpha
##    <int> <list>                <dbl> <chr>        <dbl> <dbl> <chr>        <dbl>
##  1     1 <dbl[,1] [5 x 1]>     0.749 new_basis        1     1 search_bett~   0.5
##  2     2 <dbl[,1] [5 x 1]>     0.730 random_sear~     2     1 search_bett~   0.5
##  3     3 <dbl[,1] [5 x 1]>     0.743 random_sear~     2     2 search_bett~   0.5
##  4     4 <dbl[,1] [5 x 1]>     0.736 random_sear~     2     3 search_bett~   0.5
##  5     5 <dbl[,1] [5 x 1]>     0.747 random_sear~     2     4 search_bett~   0.5
##  6     6 <dbl[,1] [5 x 1]>     0.725 random_sear~     2     5 search_bett~   0.5
##  7     7 <dbl[,1] [5 x 1]>     0.752 new_basis        2     6 search_bett~   0.5
##  8     8 <dbl[,1] [5 x 1]>     0.749 interpolati~     2     1 search_bett~  NA  
##  9     9 <dbl[,1] [5 x 1]>     0.750 interpolati~     2     2 search_bett~  NA  
## 10    10 <dbl[,1] [5 x 1]>     0.750 interpolati~     2     3 search_bett~  NA
\end{verbatim}

>>>>

@@

@@

{\centering \includegraphics[width=0.6\linewidth,height=0.3\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/tour_path_keynote/tour_path_keynote.001}

@@

@@

Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that provides five points summary of the data, however, there are two pieces of information missing from the boxplot: 1) It does not report the number of points, and 2) the position of the last basis. This could be remedied by adding more layers using the concept of grammar of graphics. A label geometry is added at the bottom of the plot to show the number of points in each iteration and a line geometry links the last basis in each iteration. Further, an option to switch between displaying points and boxplot geometry is helpful since point geometry can be more intuitive for the iteration with few observations. This is achieved via a \texttt{cutoff} parameter.

@@

Sometimes, we may be interested in exploring the points on the interpolation path since these are the points that will be played by the tour animation. Figure \ref{toy-interp} presents the interpolation of three different tour paths. The upper plot shows a desirable interpolation in each iteration with the index value being progressively and monotonically increasing. While in the middle plot, the increases towards the target basis is not monotonical in the last two iterations and interpolated basis with higher index value can be found on the tour path. The lower plot is constructed using \texttt{search\_better\_random}, where a basis with smaller index value has a probabilistic chance of being accepted and we can observe a much more involved pattern. In iteration three, the probabilistic acceptance produces a monotonically decreasing interpolation whilst in iteration five, six and seven, with also a probabilistic acceptance, the interpolation is now non-monotonical. When the target basis has a higher index value, iteration eight reaches this basis by first decreasing the index value. While the interpolation plot shows different possibilities of the change in index value on the tour path, the diagnosis of the validity of each pattern and the related optimisation algorithms will be postponed to Section \ref{application}.

@@

@@

@@

Animating the plots introduced above is useful, especially in the case of PCA polot since it shows the bases found by the optimiser in its natural order. In Figure \ref{toy-pca-animated}, \ldots{}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{toy-pca-aniamted}A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm.}
\end{figure}

@@

Once the final basis has been found by an algorithm, one may want to push further to investigate whether there's an even better basis in the close neighbourhood. This motivates the polish search where the final basis is supplied to a new guided tour to search for any local breakthrough.

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}

\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1.pdf}

\includegraphics{paper_files/figure-latex/unnamed-chunk-4-1.pdf}

Comparison between search\_better\_random and two different search\_better. In the trace plot, the two dashed horizontal lines are the index values of the two theoretical bases. serach\_better\_random and one of the serach\_better have found the global maximum but the other fails. The lower panel shows the PCA plot for each of the three algorithms. The two yellow dots, again, correspond to the theoretical bases with the one on the right being the global maximum. The paths of two search\_better are relatively short and simple while the search\_better\_random has a more sophisticated path. This is due to the fact that the interpolation at each iteration are longer, which shows its ability to find new bases at a distance

@@

The implementation of this projection has been divided into two pacakges, where the data collection happens when a tour is running and hence implemented in the existing \texttt{tourr} package. The diagnostics of the algorithm via plots and other auxiliary functions have been implemented in a new pacakge, \texttt{ferrn}.

@@

\begin{verbatim}
## # A tibble: 5 x 8
##      id basis             index_val info          tries  loop method       alpha
##   <int> <list>                <dbl> <chr>         <dbl> <dbl> <chr>        <dbl>
## 1     1 <dbl[,1] [5 x 1]>     0.749 new_basis         1     1 search_bett~   0.5
## 2     2 <dbl[,1] [5 x 1]>     0.730 random_search     2     1 search_bett~   0.5
## 3     3 <dbl[,1] [5 x 1]>     0.743 random_search     2     2 search_bett~   0.5
## 4     4 <dbl[,1] [5 x 1]>     0.736 random_search     2     3 search_bett~   0.5
## 5     5 <dbl[,1] [5 x 1]>     0.747 random_search     2     4 search_bett~   0.5
\end{verbatim}
